---
title: "Extras: Google BigQuery and GDELT 2.0 Global Knowledge Graph"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

## The GDELT Projcet

[The GDELT Project](http://gdeltproject.org/data.html#rawdatafiles) provides access to a massive database of news events from
around the world. It "is an open platform for research and analysis of global society and thus all datasets released by 
the GDELT Project are available for unlimited and unrestricted use for any academic, commercial, or governmental use of any kind without fee."

## Working with GDELT 2.0 through BigQuery

Google BigQuery gives you access to massive computing power

You can sign up for a free account at [cloud.google.com](cloud.google.com). There is a 
free tier allowing 1TB of processing per month. With a free account you will also get
~$300 in credit.

Let's grab 10 themes from the GDELT Global Knowledge Graph (GKG). The below query looks for themes in 
events that mention people with names like 'Kenyatta' between the dates of 1 March and 1 April 2017.

```{r}
#install.packages("bigrquery")
devtools::install_github("rstats-db/bigrquery")
library(bigrquery)

project <- "busarabigquerydemo"

sql <- "SELECT * from [gdelt-bq:gdeltv2.gkg] where DATE>20170301000000 and DATE < 20170401000000 and V2Persons like '%Kenyatta%' limit 10;"

result <- query_exec(sql, project = project)
```

```{r}
result
```


The cost of queries is $5 per terabyte, so this can get expensive quickly ([pricing](https://cloud.google.com/bigquery/pricing)).

See [http://blog.gdeltproject.org/a-compilation-of-gdelt-bigquery-demos/](http://blog.gdeltproject.org/a-compilation-of-gdelt-bigquery-demos/)
for more examples on using BigQuery to explore the GDELT. 
